---
title: "MLS Group Stage simulation"
output: html_document
date: "2022-12-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Cross conference group stage vs divided conference group stage comparison

Steps:
+ randomly assign each team a quality score, drawn from uniform distribution
+ have teams in each conference play round robin where chance of victory is proportional to quality ratio
+ select top 8 teams in each conference, divide 4 levels of seeds
+ perform two draws:
  + one cross conference
  + one where conferences are drawn into groups separately
+ measure average quality of opponents in each scenario, record seed and average opponent quality for both draw types
+ repeat above 1000 times (see part 2)

## Part 1: single run example
### assign team qualities, run regular season
```{r team_quality_matches}
# assign quality scores to each team in east and west
east<-data.frame(teams=letters[1:18],qualities=rnorm(18,mean=100,sd=30))
west<-data.frame(teams=letters[1:18],qualities=rnorm(18,mean=100,sd=30))
# perform round-robin in each tournament
round_robin<-function(conference){
  # create a list to store match results
  match_results<-list()
  for(team1 in letters[1:18]){
    for(team2 in letters[1:18]){
      if(team1!=team2){
        # get quality scores for each team
        team1_quality=conference%>%
          filter(teams==team1)%>%
          select(qualities)%>%
          unlist()%>%
          as.numeric()%>%
          max(5) # minimium win probability
        team2_quality=conference%>%
          filter(teams==team2)%>%
          select(qualities)%>%
          unlist()%>%
          as.numeric()%>%
          max(5) # minimum win prability weight
        # randomly assign 3 points to team 1 or team 2
        # result=0 means team1 wins, result=1 means team2 wins
        game_result<-sample(c(0,1,2),1,F,c(team1_quality,
                                           mean(team1_quality,team2_quality),
                                           team2_quality))
        team1_points<-0
        team2_points<-0
        if(game_result==0){
          team1_points<-3
        }else if(game_result==1){
          team1_points<-1
          team2_points<-1
        }else{
          team2_points<-3
        }
        match_frame=data.frame('home'=c(team1),
                               'away'=c(team2),
                               'home_points'<-c(team1_points),
                               'away_points'<-c(team2_points))
        match_results[[paste0(team1,team2)]]<-match_frame
      }
    }
  }
  result_frame<-do.call(rbind,match_results)
  names(result_frame)<-c("home","away","home_points","away_points")
  return(result_frame)
}
# function that calculates standings based on conference matches

east_results<-round_robin(east)
west_results<-round_robin(west)
east_results
west_results
```

### calculating standings
```{r standings, echo=FALSE}
standings<-function(conference,conference_results){
  inner_join(conference_results%>%
         group_by(home)%>%
         summarize(home_points=sum(home_points)),
       conference_results%>%
         group_by(away)%>%
         summarize(away_points=sum(away_points)),
       by=c("home"="away"))%>%
    mutate(points=home_points+away_points)%>%
    arrange(desc(points))%>%
    select("teams"=home,points)%>%
    mutate(placement=row_number(),
           seed_pot = case_when(placement<=2~"pot1",
                                placement<=4~"pot2",
                                placement<=6~"pot3",
                                placement<=8~"pot4"))%>%
    left_join(conference,by="teams")
}
east_standings<-standings(east,east_results)%>%
  mutate(conf_team=paste0('East_',teams))
west_standings<-standings(west,west_results)%>%
  mutate(conf_team=paste0('West_',teams))
east_standings
west_standings
bind_rows(east_standings%>%
            mutate(conference='East'),
          west_standings%>%
            mutate(conference='West'))%>%
  ggplot(aes(x=qualities,y=placement,color=conference))+
  geom_point()+
  labs(x="team quality",y="table place",color="conference",
       title="Team Quality vs League Table Place")+
  theme(legend.position='bottom')
```

### Perform group draws
```{r perform_draws}
# draw 1 - separate conferences
divided_draw<-function(standings){
  standings%>%
  # eliminate seeded teams
  filter(!is.na(seed_pot))%>%
  # randomize order within pots, first in pot = Group A
  mutate(draw=runif(nrow(.)))%>%
  arrange(seed_pot,draw)%>%
  mutate(group=case_when(row_number()%%2==1~'groupA',TRUE~'groupB'))
}
east_draw<-divided_draw(east_standings)
west_draw<-divided_draw(west_standings)
# draw2 - east and west combined
combined_draw<-function(east_standings,west_standings){
  bind_rows(east_standings,west_standings)%>%
  # eliminate seeded teams
  filter(!is.na(seed_pot))%>%
  # randomize order within pots, first in pot = Group A
  mutate(draw=runif(nrow(.)))%>%
  arrange(seed_pot,draw)%>%
  mutate(group=case_when(row_number()%%4==1~'groupA',
                         row_number()%%4==2~'groupB',
                         row_number()%%4==3~'groupC',
                         T~'groupD'))
}
east_west_draw<-combined_draw(east_standings,west_standings)
east_draw
west_draw
east_west_draw
```
### measure opponent quality in group for each team
```{r opponent_qual}
# function to find opponent quality per seed for given draw
find_opponent_qual<-function(draw){
  group_qual<-draw%>%
    group_by(group)%>%
    summarize(group_quality=sum(qualities))%>%
    ungroup()
  opponent_qual<-draw%>%
    left_join(group_qual,by="group")%>%
    # total quality of teams in group minus given team's quality
    mutate(opponent_qual=group_quality-qualities)%>%
    select(placement,conf_team,group_quality,qualities,opponent_qual)
  return(opponent_qual)
}
east_qual<-find_opponent_qual(east_draw)
west_qual<-find_opponent_qual(west_draw)
combined_qual<-find_opponent_qual(east_west_draw)

opponent_quals<-bind_rows(
  east_qual%>%
    mutate(draw_type='separate'),
  west_qual%>%
    mutate(draw_type='separate'),
  combined_qual%>%
    mutate(draw_type='combined')
)
opponent_quals%>%
  ggplot(aes(x=placement,y=opponent_qual,color=draw_type))+
  geom_point(position = position_jitter(w = 0.05, h = 0))+
  geom_smooth(method='lm')+
  theme(legend.position="bottom")+
  labs(y='sum of group stage opponents quality scores',
       x='conference finish place',
       title='single simulation run')
```

## Part 2: results of 100 runs

```{r monte_carlo, cache=TRUE}
n_sim=100
sim_results<-list()
league_standings<-list()
for(sim in 1:n_sim){
  # assign quality scores to each team in east and west
  east<-data.frame(teams=letters[1:18],qualities=rnorm(18,mean=100,sd=30))
  west<-data.frame(teams=letters[1:18],qualities=rnorm(18,mean=100,sd=30))
  # run round-robin regular season
  east_results<-round_robin(east)
  west_results<-round_robin(west)
  # calculate standings
  east_standings<-standings(east,east_results)%>%
    mutate(conf_team=paste0('East_',teams))
  west_standings<-standings(west,west_results)%>%
    mutate(conf_team=paste0('West_',teams))
  # append east and west standings to league_standings so we can verify quality correlates with placement
  league_standings[[paste0(sim)]]<-bind_rows(east_standings%>%
            mutate(conference='East'),
          west_standings%>%
            mutate(conference='West'))
  # perform draws
  east_draw<-divided_draw(east_standings)
  west_draw<-divided_draw(west_standings)
  east_west_draw<-combined_draw(east_standings,west_standings)
  # calculate opponent qualities
  east_qual<-find_opponent_qual(east_draw)
west_qual<-find_opponent_qual(west_draw)
combined_qual<-find_opponent_qual(east_west_draw)
opponent_quals<-bind_rows(
  east_qual%>%
    mutate(draw_type='separate'),
  west_qual%>%
    mutate(draw_type='separate'),
  combined_qual%>%
    mutate(draw_type='combined')
)
  # store results
  sim_results[[paste0(sim)]]<-opponent_quals
}
```

```{r plot_monte}
sim_results%>%
  bind_rows()%>%
  mutate(seed=case_when(placement<=2~1,
                        placement<=4~2,
                        placement<=6~3,
                        placement<=8~4))%>%
  ggplot(aes(x=placement,y=opponent_qual,color=draw_type))+
  geom_point(position = position_jitter(w = 0.2, h = 0),alpha=.3)+
  geom_smooth(method='lm')+
  theme(legend.position="bottom")+
  labs(y='sum of group stage opponents quality scores',
       x='conference finish place',
       title=paste0(n_sim,' trial simulation results'))
league_standings%>%
  bind_rows()%>%
  ggplot(aes(x=qualities,y=placement,color=conference))+
  geom_point()+
  labs(x="team quality",y="table place",color="conference",
       title="Team Quality vs League Table Place")+
  theme(legend.position='bottom')+
  scale_color_discrete(palette('dark2'))
  
```